GPU-based Adaptive Beamforming for Active Sonar Imaging

### 2012-05-08 ECUA2012/gpu_capon_ecua2012.tex
la2pdf commit - 2012:05:08	10:14:07	(Tue 8.May)
la2pdf commit - 2012:05:08	10:46:47	(Tue 8.May)

The Minimum Variance (MV) adaptive beamformer has recently been introduced in active sonar imaging. While displaying great potential for creating high quality images, it is computationally intensive and relies on robustification. These are important reasons why the MV beamformer is not in widespread use today.

@@

The Minimum Variance Distortionless Response (MVDR, or Capon) beamformer is one such technique. While ensuring unit gain in the look direction, it computes the set of weights that minimizes the power accumulated by the array from other directions. Its ability to suppress interference power and improve image quality in active sonar systems has been demonstrated in e.g.~\cite{Blomberg2012a,Blomberg2011,Dursun2009,Lo2004}.

@@

\emph{Spatial averaging} is achieved by grouping the array channels into $M-L+1$ subarrays of length $L$, where all but one element overlap. If we let $x_l[n]$ represent the datavector from subarray $l$,
\begin{gather}
\x_l[n] = \bmat{x^\Delta_l[n] & x^\Delta_{l+1}[n] & \dots & x^\Delta_{l+L-1}[n]}\T,
\end{gather}
then the ``subarray averaged'' sample covariance matrix, $\eR_\text{s}$, is given as:
\begin{gather}
\eR_\text{s}[n] = \frac{1}{M-L+1} \sumb{l=0}{M-L} \x_l[n]\x_l\H[n].\label{spatialR}
\end{gather}

\emph{Temporal averaging} can now be introduced by taking the average of $2N_\text{avg}+1$ temporal samples of $\eR_\text{s}[n]$. Let the result be called $\eR_\text{t,s}[n]$,
\begin{gather}
\eR_\text{t,s}[n] = \frac{1}{2N_\text{avg}+1} \sumb{n'=n-N_\text{avg}}{n+N_\text{avg}} \eR_\text{s}[n'].
\end{gather}
Finally, to ensure a well conditioned sample covariance matrix, we add a fraction $d$ of the total power of $\eR_\text{t,s}[n]$ to its diagonal\cite{Synnevag2007}:
\begin{align}
\eR[n] = \eR_\text{t,s}[n] + \I \cdot \frac{d}{L} \cdot \tr\{\eR_\text{t,s}[n]\},\label{finalR}
\end{align}
where $\I$ is an identity matrix and $\tr\{\cdot\}$ means matrix trace operation.

@@

This suggests that the algorithm should be decomposed into as many lightweight threads as possible, which are both computationally intensive and light on the memory access and consumption. However, for a typical array configuration, using a single thread per pixel is not lightweight enough. Therefore, the methods presented section \ref{methods} had to be decomposed further. In particular, the solutions we came up with for the different steps in the algorithm, listed in their natural order of execution, was:
\begin{enumerate}
\item \emph{Computing the spatial covariance matrix} $\eR[n]$ (\ref{spatialR}-\ref{finalR}). A group of threads are created that slide along the diagonals of $\eR$. In this way we exploited the fact that entries on the diagonals overlap across subarrays and time, and keep the numbers of both data reads and writes at a minimum.
%Finished threads can also wrap around and start on diagonals on the lower half triangular. 
% In this way we process one row of $\eR$ per kernel iteration, and manage to keep the numbers of both data reads and writes at a minimum.
\item \emph{Computing} $\eRi[n]\1$ (from (\ref{weights})). While intuition may suggest that this step is carried out by inverting $\eR$, it is better to solve the linear equation $\R\beta = \1$ for $\beta$ instead, which gives us $\beta = \eRi\1$ directly. We have tested various solvers for this task, both in-house and proprietary implementations, and achieved the best performance by using an unofficial batch solver from nVidia. Inverting $\eR$ is by far the most computationally intensive task, and a key area of focus for further improvements.
\item \emph{Computing the beamformer output} $z[n]$ by substituting $\beta = \eRi[n]\1$ into (\ref{weights}), and (\ref{weights}) into (\ref{z}). A group of $L$ threads per pixel was used to reduce local memory pressure and to obtain coalesced reads and a minimum of writes. When applying weights, there are two approaches from an implementation point of view. The subarray data can, as here, be reduced to coincide in length with the subarray weights, or the weights can be extended to $M$ in size and applied directly on $\x[n]$ and summed.  
\end{enumerate}

@@

\newpage
Interesting to note is also that solving $\eRi\1$ remains the main bottleneck in the \gls{GPU} implementation, and it gets even worse for larger $L$'s. However, this is a general Gauss Jordan solver which does not exploit symmetry properties of $\eR$, so creating e.g.\ a batch based Cholesky solver should improve the runtimes further. Both the C and Matlab implementation in this benchmark take this symmetry into account, which partially explains why the complexity curves of these implementations are different from that of the \gls{GPU} implementation.

>>>>

The Minimum Variance Distortionless Response (MVDR) adaptive beamformer has recently been introduced in active sonar imaging. While displaying great potential for creating high quality images, it is computationally intensive and relies on robustification. These are important reasons why the MV beamformer is not in widespread use today.

@@

The MVDR adaptive beamformer (also known as the Minimum Variance or Capon beamformer) is one such technique. While ensuring unit gain in the look direction, it computes the set of weights that minimizes the power accumulated by the array from other directions. Its ability to suppress interference power and improve image quality in active sonar systems has been demonstrated in e.g.~\cite{Blomberg2012a,Blomberg2011,Dursun2009,Lo2004}.

@@

\emph{Spatial averaging} is achieved by grouping the array channels into $M-L+1$ subarrays of length $L$, where all but one element overlap. If we let $x_l[n]$ represent the datavector from subarray $l$,
\begin{gather}
\x_l[n] = \bmat{x^\Delta_l[n] & x^\Delta_{l+1}[n] & \dots & x^\Delta_{l+L-1}[n]}\T,
\end{gather}
then the ``subarray averaged'' sample covariance matrix, $\eR_\text{s}$, is given as:
\begin{gather}
\eR_\text{s}[n] = \frac{1}{M-L+1} \sumb{l=0}{M-L} \x_l[n]\x_l\H[n].\label{spatialR}
\end{gather}
\emph{Temporal averaging} can now be introduced by taking the average of $2N_\text{avg}+1$ temporal samples of $\eR_\text{s}[n]$. Let the result be called $\eR_\text{t,s}[n]$,
\begin{gather}
\eR_\text{t,s}[n] = \frac{1}{2N_\text{avg}+1} \sumb{n'=n-N_\text{avg}}{n+N_\text{avg}} \eR_\text{s}[n'].
\end{gather}
Finally, to ensure a well conditioned sample covariance matrix, we add a fraction $d$ of the total power of $\eR_\text{t,s}[n]$ to its diagonal\cite{Synnevag2007}:
\begin{align}
\eR[n] = \eR_\text{t,s}[n] + \I \cdot \frac{d}{L} \cdot \tr\{\eR_\text{t,s}[n]\},\label{finalR}
\end{align}
where $\I$ is an identity matrix and $\tr\{\cdot\}$ means matrix trace operation.

@@

This suggests that the algorithm should be decomposed into as many lightweight threads as possible, which are both computationally intensive and light on the memory access and consumption. However, for a typical array configuration, using a single thread per pixel is not lightweight enough. Therefore, the methods presented section \ref{methods} had to be decomposed further. In particular, the solutions we came up with for the different steps in the algorithm, listed in their natural order of execution, was:
\begin{enumerate}
\item \emph{Computing the spatial covariance matrix} $\eR[n]$ (\ref{spatialR}-\ref{finalR}). A group of threads are created that slide along the diagonals of $\eR$. In this way we exploited the fact that entries on the diagonals overlap across subarrays and time, and keep the numbers of both data reads and writes at a minimum.
%Finished threads can also wrap around and start on diagonals on the lower half triangular. 
% In this way we process one row of $\eR$ per kernel iteration, and manage to keep the numbers of both data reads and writes at a minimum.
\item \emph{Computing} $\eRi[n]\1$ (from (\ref{weights})). While intuition may suggest that this step is carried out by inverting $\eR$, it is better to solve the linear equation $\R\beta = \1$ for $\beta$ instead, which gives us $\beta = \eRi\1$ directly. We have tested various solvers for this task, both in-house and proprietary implementations, and achieved the best performance by using an unofficial batch solver from nVidia. 
\item \emph{Computing the beamformer output} $z[n]$ by substituting $\beta = \eRi[n]\1$ into (\ref{weights}), and (\ref{weights}) into (\ref{z}). A group of $L$ threads per pixel was used to reduce local memory pressure and to obtain coalesced reads and a minimum of writes. When applying weights, there are two approaches from an implementation point of view. The subarray data can, as here, be reduced to coincide in length with the subarray weights, or the weights can be extended to $M$ in size and applied directly on $\x[n]$ and summed.  
\end{enumerate}

@@

\newpage
Interesting to note is also that solving $\eRi\1$ remains the main bottleneck in the \gls{GPU} implementation, and the complexity depends on the subarray size $L$'s. This means the overall complexity of the \gls{MVDR} beamformer is actually reduced when we apply subarray averaging. However, this is a general Gauss Jordan solver which does not exploit symmetry properties of $\eR$, so creating e.g.\ a batch based Cholesky solver should improve the runtimes further. Both the C and Matlab implementation in this benchmark take this symmetry into account, which partially explains why the complexity curves of these implementations are different from that of the \gls{GPU} implementation.
